### 2017-01-22 manuscript.rmd
added ascertainment bias explorations and ortholog conjecture text

A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the statistical challenges and opportunities of comparisons across species [@Felsenstein:1985ua; Grafen:1989um; Chamberlain:2012vx; Garamszegi:2014ew; Pagel:1999gc]. These have largely been applied to the evolution of morphological and ecological traits, but are just as relevant to CFG. Most CFG studies have abstained from phylogenetic approaches and instead heavily rely on pairwise comparisons (Fig. 1A) that do not account for the fact that evolutionary relationships explain much of the structure of variation across species. This leaves CFG studies susceptible to serious statistical errors and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.

@@

@@

```{r preliminaries, echo=FALSE, message=F, warning=F}
	library( treeio )
	library( tidyverse )
	library( magrittr )
	library( stringr )
	library( parallel )
	library( ggtree )
	library( ape )
	library( digest )
	library( hutan )

@@

	# These are from the KRR code
	calibration_times = data.frame(
		age.min =
			c(20, 92, 167, 8, 42, 74, 296, 535, 104, 937, 29, 722, 441, 65, 15, 1215, 
				414, 371, 162, 25, 74, 77, 86), 
		clade = c("Hominoidea", "Euarchontoglires", "Mammalia", "Homininae", 
							"Simiiformes", "Primates", "Amniota", "Vertebrata", "Eutheria", 
							"Bilateria", "Catarrhini", "Chordata", "Euteleostomi", 
							"Haplorrhini", "Hominidae", "Opisthokonta", "Sarcopterygii", 
							"Tetrapoda", "Theria", "Murinae", "Sciurognathi", "Rodentia", "Glires"),
		stringsAsFactors=FALSE
	)

@@

	gene_trees_sub = mclapply( gene_trees_annotated, drop_empty_tips, mc.cores=cores )
	gene_trees_sub = gene_trees_sub[ ! is.na( gene_trees_sub ) ]
#	gene_trees_sub = gene_trees_sub[ lapply(gene_trees_sub, class) == "treedata" ]

@@

@@

	nodes_contrast %>%
		ggplot( aes( x=pic, y=..density.., col=D ) ) + 
        geom_freqpoly( binwidth=.001, position="identity" ) +
        xlab("Phylogenetic independent \ncontrasts (PIC)") +
				xlim(0,0.075)

	nodes_contrast %>%
		filter( pic < 0.2 ) %>%
		ggplot( aes( x=node_age, y=pic, col=D ) ) +
		geom_point( alpha=0.2) +
		xlim(0,1000)

	wilcox.test( nodes_contrast %>% filter(D=="Y") %>% .$pic, nodes_contrast %>% filter(D=="N") %>% .$pic, alternative="greater" )

@@

@@

@@

Based on their scripts from: https://figshare.com/articles/Tissue-specificity_of_gene_expression_diverges_slowly_between_orthologs_and_rapidly_between_paralogs/3493010/2

To replciate the published analysis, `cd` to `Data_used/Expression_data`, launch `R`, and run:

    source( "../Rscript.R" )

After running the analysis above, also obtain the `Compara.75.protein.nh.emf` gene trees from ftp://ftp.ensembl.org/pub/release-75/emf/ensembl-compara/homologies/ and place them in the same directory as this file.

Thanks to

>>>>

A rich theoretical and statistical methodology of phylogenetic comparative methods have been developed over the last three decades to address the challenges and opportunities of comparisons across species [@Felsenstein:1985ua; Grafen:1989um; Chamberlain:2012vx; Garamszegi:2014ew; Pagel:1999gc]. These have largely been applied to the evolution of morphological and ecological traits, but are just as relevant to CFG. Most CFG studies have abstained from phylogenetic approaches and instead heavily rely on pairwise comparisons (Fig. 1A) that do not account for the fact that evolutionary relationships explain much of the structure of variation across species. This leaves CFG studies susceptible to serious statistical errors and is a missed opportunity to ask questions that are only accessible in an explicit phylogenetic context.

@@

The KMRR considers several publicly available datasets. Here we consider their conclusions in the context of Brawand *et al.* 2011 [Brawand:2011du], one of the studies they considered. This analysis of XX different organs across XX mammal species.

For each internal node in each gene tree (Figure 1b), we then measured the phylogenetically independent contrast (PIC) in the expression across this node [@Felsenstein:1985ua]. This is the difference in values of descendent nodes (i.e., the evulutionary change along the branches that descend from the node) scaled by the expected variance (determined largely by the lengths of these branches). These contrasts were then annotated by whether they are made across branches that descend from a speciation or duplication event. The original description of independent contrasts [@Felsenstein:1985ua] focused on assessing covariance between changes in two traits on a tree. Our use of it to look for differences in evolutionary changes of one trait (differential expression) between two categories of nodes (speciation and duplication) on a tree is a bit different than the usual application of the method.

![Figure Expectations](figure_expectations.svg?raw=true)

> Figure XXExpectations | (a) Null hypothesis. (b) Under the ortholog conjecture.

To test the ortholog conjecture it is essental to have a null hypothesis that results in a distinct prediction from the ortholog conjecture. A suitable null hypothesis is that there is no difference in evolutionary change in differential expression along branches that descend from speciation or duplication events (Figure 1b). Under this hypothesis, we would predict that contrasts across speciation nodes and duplication nodes are drawn from the same distribution (Figure XXEXPECTATIONS a). Under the alternative hypothesis specified by the ortholog conjecture, that there is a higher rate of change following duplication events than speciation events, we would expect to see the distribution of duplication contrasts shifted to higher values relative to the speciation contrasts (Figure XXEXPECTATIONS b).

@@

```{r preliminaries, echo=FALSE, message=F, warning=F}
	library( treeio )
	library( tidyverse )
	library( magrittr )
	library( stringr )
	library( parallel )
	library( ggtree )
	library( ape )
	library( digest )
	library( hutan )
	library( gridExtra )

@@

	# These are from the KMRR code
	calibration_times = data.frame(
		age.min =
			c(20, 92, 167, 8, 42, 74, 296, 535, 104, 937, 29, 722, 441, 65, 15, 1215, 
				414, 371, 162, 25, 74, 77, 86), 
		clade = c("Hominoidea", "Euarchontoglires", "Mammalia", "Homininae", 
							"Simiiformes", "Primates", "Amniota", "Vertebrata", "Eutheria", 
							"Bilateria", "Catarrhini", "Chordata", "Euteleostomi", 
							"Haplorrhini", "Hominidae", "Opisthokonta", "Sarcopterygii", 
							"Tetrapoda", "Theria", "Murinae", "Sciurognathi", "Rodentia", "Glires"),
		stringsAsFactors=FALSE
	)

@@

	gene_trees_sub = mclapply( gene_trees_annotated, drop_empty_tips, mc.cores=cores )
	gene_trees_sub = gene_trees_sub[ ! is.na( gene_trees_sub ) ]

@@

	all_nodes$Event = NA

	all_nodes$Event[ all_nodes$D == "Y" ] = "Duplication"
	all_nodes$Event[ all_nodes$D == "N" ] = "Speciation"

@@

	nodes_contrast %>%
		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
        geom_freqpoly( binwidth=.001, position="identity" ) +
        xlab("Phylogenetic independent \ncontrasts (PIC)") +
				xlim(0,0.075) + 
				theme_classic()

	wilcox_test_result = wilcox.test( nodes_contrast %>% filter(D=="Y") %>% .$pic, nodes_contrast %>% filter(D=="N") %>% .$pic, alternative="greater" )

@@

> Figure XXKMRR | Density plot of the magnitude of phylogenetic independent contrast values following duplication and speciation events. These contrasts are not larger following duplication events, as predicted by the ortholog conjecture.

@@

Given the ortholog conjecture, we would expect to see larger expression contrasts following duplication events than speciation events. Instead,we find no such distribution (Figure XXKMRR). The Wilcox rank test does not reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (p value = `r wilcox_test_result$p.value`). Our results therefore find no evidence for the ortholog conjecture in this system.

We next examined the possibility that ascertainment biases were differentially impacting the inference of evolutionary change in differential expression following duplication and speciation events. We focused on two possible sources of bias - node depth and branch length.

While the age of speciation nodes is constrained, duplication nodes can be much older and therefore have a wider range of ages (Supplementary Figure XXAscertainment a). Biological reasons for this include the fact that duplication events can be older than the root of the species tree, but speciation events cannot. There are also technical factors that can lead to an excess of duplication events deeper in the tree, including the impact of assembly errors that lead to the overestimation of deep duplications [@Hahn:2007cj]. If independent contrast values also tended to to be lower at greater node depth, it could interact with the preponderance of duplications at greater depth to create a pattern of lower contrasts following duplication events. When we remove duplication nodes that are older than the oldest speciation node, our results are unchanged and we cannot reject the null hypothesis that the rate of evolution following duplications is the same as or less than the rate following speciation (Supplementary Figure XXAscertainment b).

The independent contrast across a node is the amount of change observed between the daughter notes, scaled by the expected variance [Felsenstein:1985ua]. The expected variance is principaly determined by the lengths of the branches leading from the node to these daughters. The same observed difference between daughter nodes will lead to a larger contrast the shorter the branches are to these nodes, as this indicates a greater rate of evolutionary change. The expected variance of contrasts for speciation nodes is constrained by the branch lengths on the species tree, but the expected variance of contrasts for duplications can range much wider (Figure XXAscertainment c).

The fact that the expected variance of speciation contrasts is a subset of the range of the expected variance of duplication contrasts could lead to ascertainment biases. We therefore also examine only the contrasts that fell within in the range of expected variance seen for speciation contrasts, excluding duplication contrasts that fall outside of this range. This reanalysys does not reject the ortholog conjecture either (Figure XXAscertainment d).

The widespread invokation of the ortholog conjecture follows from the expectation that it is strong, widely distributed phenomenon. Our failure to detect a signature consistent with the ortholog conjecture suggests that it may not be, and that it should not be assumed in systems where it has not been explicitly tested.

### Comparative methods in functional genomics

### Ortholog conjecture

The ortholog conjecture [@Nehrt:2011bm] is the proposition that orthologs (genes that diverged from each other due to a speciation event) have more similar attributes than do paralogs (genes that diverged from each other due to a gene duplication event). It has important biological and technical implications. It shapes our understanding of the functional diversity of gene families. It is used to relate findings from well-studied genes to related genes that haven't been investigated in detail. It can be applied to any trait of genes, from gene seqeunce to biochemical properties to expression. While the ortholog conjecture describes a specific pattern of functional diversity across genes, it can also be considered a hypothesis about the process of evolution. Specifically, it postulates that there is greater evolutionary change in gene attributes following a duplication event than a speciation event.

Despite its importance, there have been relatively few tests of the ortolog conjecture. Previous work has shown that ontology annotations are not sufficient to test the ortholog conjecture [@Chen:2012jz; @Thomas:2012hz]. Analyses of domain structure were consistent with ortholog conjecture [@Forslund:2011cw]. There have been few tests of the ortholog conjecture with regards to gene expression [@Chen:2012jz].

Our reanalysis of KMRR using phylogenetic methods suggests that the ortholog conjectures is not a dominant pattern that is central to explaining the evolution of phenotypic diversity in gene families. This has impotant biological implications. It suggests that the mechanism of gene divergence (speciation versus duplication) may not have as strong an impact on phenotypic divergence as thought. It also has technical implications. Rather than focus on whether genes are orthologs or paralogs when attempting to predict function, it may be more effective to simply focus on how closely related or distantly related they are. Closely related paralogs, for example, may tend to have more similar phenotypes than more distantly related orthologs.

All files needed to reexecute the anlyses presented in this document are available at https://github.com/caseywdunn/comparative_expression_2017. The most recent commit at the time of the analysis presented here was `r system("git log | head -n 1", intern=TRUE) %>% str_replace("commit ", "")`.

The KMRR study [@KryuchkovaMostacci:2016iw] followed excellent practices in reproducibility. They posted all data and code needed to re-execute their analyses at figshare:  https://figshare.com/articles/Tissue-specificity_of_gene_expression_diverges_slowly_between_orthologs_and_rapidly_between_paralogs/3493010/2 . We slightly altered their `Rscript.R` to simplify file paths. This midified script and their data files are available in the gihub respository for this paper, as are the intermediate files that were generated by their analysis script that we used in our own analyses. We also obtained the `Compara.75.protein.nh.emf` gene trees [@Herrero:2016jj] from ftp://ftp.ensembl.org/pub/release-75/emf/ensembl-compara/homologies/ and place them in the same directory as this file.

Each of these gene trees includes includes homologous sequences derived from a common ancestral sequence. The tip nodes of a gene tree can include both orthogs and paralogs, and each internal node is an infered speciation or duplication event. Prior to phylogenetic analyses of gene expression, each tree was first pruned to include only tips for which expression data were available. The speciation nodes in the tree were then time calibrated so that branch lengths were consistent across trees. We then calculated the magnitude of the independent contrast for differential expression at each internal node in each of these trees. The independent contrast is the amount of change in expression observed along the branches that arise from the node, scaled by the expected variance [Felsenstein:1985ua]. The expected variance is determined mostly by the length of the branches, but can also be influenced by uncertainty regarding the reconstruction of differential expression at internal nodes. We took the absolute value of these contrasts, since the sign depends only on the arbitrary ordering of the nodes in the trees.

Thanks to XX. This work was supported by the National Science Foundation (DEB-1256695 and the Waterman Award).

### KMRR Analyses

```{r filtered, cache=TRUE, echo=FALSE, fig.cap="XX_PIC_VAR", warning=FALSE, message=FALSE}

	# Examine by node age

  g_age =
  nodes_contrast %>%
    ggplot( aes( x=node_age, y=pic, col=Event ) ) + 
		  geom_point( alpha=0.2 ) + 
		  xlab( "Node age (millions of years)" ) + 
		  ylab("Phylogenetic independent contrasts (PIC)\n of differential expression") +
		  ylim(0,0.3) +
			xlim(0,1000) +
		  ggtitle( "(a)" ) +
		  theme_classic()	
		  #theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) )

	max_speciation_age = 
    nodes_contrast %>% 
    filter( D=="N" ) %>%
    select( node_age ) %>%
    max( na.rm=TRUE )

	nodes_contrast_filtered_age = 
		nodes_contrast %>% 
		filter( node_age <= max_speciation_age )

	wilcox_filtered_age = 
		wilcox.test( 
			nodes_contrast_filtered_age %>% filter(D=="Y") %>% .$pic, 
			nodes_contrast_filtered_age %>% filter(D=="N") %>% .$pic, 
			alternative="greater" 
		)$p.value

	label_age = str_c( "Wilcoxon p=", wilcox_filtered_age )

	g_density_filtered_age =   
		nodes_contrast_filtered_age %>%
		ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
      geom_freqpoly( binwidth=.001, position="identity" ) +
      xlab("Phylogenetic independent \ncontrasts (PIC)") +
			ylab("Density") +
			xlim(0,0.075) + 
			ggtitle( "(b)" ) +
			annotate("text", x = Inf, y = Inf, hjust="right", vjust="top", label=label_age ) +
			theme_classic()

	# Examine by expected variance (branch length)
  g_var_exp =
    nodes_contrast %>%
      ggplot( aes( x=var_exp, y=pic, col=Event ) ) + 
      geom_point( alpha=0.2 ) + 
      xlab( "Expected variance, \n depends mostly on branch length" ) + 
      ylab("Phylogenetic independent contrasts (PIC)\n of differential expression") +
      ylim(0,0.3) +
			xlim(0,2000) +
      ggtitle( "(c)" ) +
      theme_classic()	
      # theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) )

  var_exp_max = 
    nodes_contrast %>% 
    filter( D=="N" ) %>%
    select( var_exp ) %>%
    max( na.rm=TRUE )

  var_exp_min = 
    nodes_contrast %>% 
    filter( D=="N" ) %>%
    select( var_exp ) %>%
    min( na.rm=TRUE )

  nodes_contrast_filtered_var = 
    nodes_contrast %>% 
      filter ( var_exp >=var_exp_min & var_exp <=var_exp_max )

	wilcox_filtered_var = 
	wilcox.test( 
		nodes_contrast_filtered_var %>% filter(D=="Y") %>% .$pic, 
		nodes_contrast_filtered_var %>% filter(D=="N") %>% .$pic, 
		alternative="greater" 
	)$p.value

	label_var = str_c( "Wilcoxon p=", wilcox_filtered_var )

  g_density_filtered_var =   
    nodes_contrast_filtered_var  %>% 
    ggplot( aes( x=pic, y=..density.., col=Event ) ) + 
      geom_freqpoly( binwidth=.001, position="identity" ) + 
      ggtitle( "(d)" ) +
      theme( legend.justification=c( 1, 1 ), legend.position=c( 1, 1 ) ) +
      xlim(0,0.075) + 
			theme_classic() +
  		annotate("text", x = Inf, y = Inf, hjust="right", vjust="top", label=label_var ) +
      xlab("Phylogenetic independent contrasts (PIC)\n of differential expression") +
  		ylab("Density")

  grid.arrange(
  	g_age,
  	g_density_filtered_age,
		g_var_exp,
		g_density_filtered_var,
		ncol = 2
	)
```

> Supplementary Figure XXAscertainment | Investigation of possible assertainment biases. (a) Magnitude of independent contrasts ploted against node age. Speciation nodes are calibrated to particular times, whereas duplication nodes have a wider range. (b) Density plot of contrasts for only the nodes that have an age less than or equal to the maximum age of speciation nodes. (c) Magnitude of independent contrasts ploted against expected variance, which is largely determined by branch lengths. Contrasts for speciation nodes have a narrower range of expected variance than do contrasts for duplication nodes. (d) Density plot of contrasts for only the nodes that have expected variance within the range of contrasts across speciation nodes.

@@

> Supplementary Figure XXPairwise | Pairwise comparisons
